Measurement of total hadronic differential cross sections in the LArIAT experiment

### 2018-04-30 Chapter-2/2-LArTPCs.tex
s

Historically, the active medium in ionization detectors has been in the gaseous form. Carlo Rubbia first proposed the use of a Liquid Argon TPC for a neutrino experiment, ICARUS \cite{Rubbia:1977zz}, in 1977.  Using nobles elements in the liquid form for neutrino detectors is advantageous for several reasons.  The density of liquids is $\sim$1000 times greater than gases, augmenting the number of targets for neutrino's interaction in the same volume, in a effort to balance the smallness of neutrino cross section. Since the energy loss of charged particle is proportional to the target material density, as shown in the Bethe-Block equation (eq. \ref{eq:BB}), the increased density reflects into a proportionally higher energy loss, enhancing the calorimetry capability of detectors with a liquid active medium. Additionally, the ionization energy of liquids is smaller than gasses by the order of tens of eV. Thus, at the passage of charged particles, liquid generally produce more ionization electrons than gas for the same deposited energy and force the particles to deposit more energy in a shorter range. The downside of using noble liquid elements in experiments is that they require expensive cryogenic systems to cool the gas until it transitions to its the liquid form.
The properties of liquid argon in comparison liquid xenon -- a popular choice for dark matter and neutrinoless double beta decay detectors -- are summarized in table \ref{tab:properties}.  Albeit xenon would be more desirable than argon given some superior properties such as lower ionization energy and higher density and light yield, argon relative abundance abates the cost of argon compared to xenon, making argon a more viable choice for the construction of kilo-ton scale neutrino detectors.

@@

LArTPCs are some times referred as to ``electronic" bubble-chambers, for the similarity in the tracking and energy resolution which is coupled with an electronic readout of the imaging information in LArTPCs. Compared to these historic detectors however, LArTPC bestow tridimensional tracking and a self triggering mechanism provided by the scintillation light in the noble gas.  An event display of a $\nu_\mu$ CC interaction candidate in the MicroBooNE detector is shown in picture \ref{fig:NuEvd} to display the level of spatial details these detectors a capable of; the color scale of the image is proportional to the energy deposited, hinting to the calorimetry capabilities of the detectors.
\begin{figure}[hbpt]
\centering
\includegraphics[width=\textwidth]{Chapter-2/Images/run3469_subrun1064_event53223_col.pdf}
\caption{Event display of a $\nu_\mu$ CC interaction candidate in the MicroBooNE detector.}
\label{fig:NuEvd}
\end{figure}

@@

%The time projection chamber (TPC) was invented by Nygren in 1974 [91]. The proposal to implement a liquid argon TPC (LArTPC) for neutrino physics was made by Rubbia in 1977 [92], and the concept was implemented by the ICARUS collaboration [93].

To the bare bones, a LArTPC is a bulk of liquid argon sandwiched in a flat capacitor, equipped with a light collection system. A uniform electric field of the order of 500 V/cm is maintained constant between the faces of the capacitor. The anode is sensitive to ionization charge and it is usually made of two or more planes segmented into several hundreds parallel sense wires a few millimeters apart; different geometries for the anode segmentation are under study \cite{1748-0221-8-07-P07002}.

Argon ionization and scintillation are the processes leveraged to detect particles in the LArTPC active volume.  When a ionizing radiation traverses the argon active volume it leaves a trail of ionization electrons along its trajectory and it excites the argon producing of scintillation light -- details on the production and detection of ionization charge and scintillation light are provided in \ref{sec:light} and \ref{sec:light} respectively. The optical the detector sees the argon scintillation light in matters of nanoseconds. This flash of light determines the start time of an event in the chamber, $t_0$. The uniform electric field drifts the ionization  electrons from the production point towards the anode in order of hundreds of microseconds or more depending on the chamber dimensions\footnote{The ionized argon also drifts, but  in the opposite directions compared to the electrons. Since the drift time is proportional to the particle mass,  the ions' drift time is much longer than the electrons'.  Ionized argon is collected on the cathode which is not instrumented, so it is not used to infer information about the interactions in the chamber.}. The anode sense wires see either an induced current by the drifting charge (on induction planes) or an injection of the ionization charge (collection plane).    An appropriate choice of the voltage bias on each wire plane assures ideal charge transparency, so that all the ionization charge is collected on the collection plane and none on the induction planes.

@@

Since the charge deposited by the ionizing radiation is proportional to the deposited energy and the charge collected on the sense plane is a function of the deposited charge, LArTPC allow the measurement of the energy deposit in the active volume. Effects due to the presence of free charge and impurities in the active volume, such as a finite electron lifetime, recombination and space charge, complicate the relationship between deposited and collected charge affecting the measurement of the particle's energy, as described in the next section.

@@

Excitation and ionization of the detector medium occur in similar amounts. Since the ionizing collisions occur randomly, we can parametrize their number $k$ in a segment of length $s$ along the track  with a Poissonian function
\begin{equation}
P(k) = \frac{s^k}{k! \lambda^k} e^{-s/ \lambda }, 
\end{equation}
where $\lambda = 1/N_{e}\sigma_i$, with $N_{e}$ being the electron density of $\sigma_i$ the ionization cross-section per electron.  About 66\% of the ionizing collisions in Argon produce only a single electron/ion pair \cite{0034-4885-73-11-116201}; in the other cases, the transferred kinetic energy is enough for the primary electron to liberate one or more secondary electrons, which usually stay close to the original pair.  
Occasionally, electrons can receive enough energy to be ejected with high energy, forming so-called ``$\delta$-ray": a detectable ionization short track off the particle trajectory, as shown in figure \ref{fig:delta}. 
The average number of $\delta$-ray  with energy E$>$E$_0$ per cm follows the empirical form
\begin{equation}
P(E>E_0) \sim \frac{y}{\beta^2 E_0},
\end{equation}
where $y$ is an empirical factor depending on the medium (0.114 for  gaseous Ar), and $\beta$ is $v/c$.

@@

\subsubsection{Purity \& Electron Life Time }
The presence of electronegative contaminants in liquid argon, such as oxygen and water, is particularly
pernicious, since these molecules quench the charge produced by the ionizing radiation.  Thus, amount of charge per unit of length $dQ/dx$ collected on the collection plane depends on the charge's production point in the detector: ionization produced  close to the cathode will see more impurities along its journey to the collection plane than ionization produced close to the anode, resulting in greater attenuation of its charge. As a result,  the amount of charge collected on the sense wires as a function of the traveled distance follows an exponential decay trend. The traveled distance is generally measured in terms of drift time and the  characteristic time constant of the exponential decay is called electron lifetime $\tau_e$. Figure \ref{fig:Elifetime} shows the typical life time for LArIAT data. The procedure to measure the electron lifetime in LArIAT is outlined in \cite{LArIATLifeTime}. LArIAT small drift distance (47 cm) allows for a relatively short electron life time. The life time for bigger detectors such as MicroBooNE, whose drift distance is 2.5 m, needs to be of the order of tens of milliseconds to allow charge collection usable for physics analyses. Energy reconstruction in LArTPC applies a correction for the finite lifetime to calibrate the detector calorimetric response; details for LArIAT are provided in Section \ref{ch:energyCalibration}.

@@

Models for a theoretical descriptions of recombination based on the Birks model and the Box model are provided in \cite{0370-1298-64-10-303} and  \cite{PhysRevA.36.614}, respectively. The Birks model assumes a gaussian spatial distribution around the particle trajectory during the entire recombination phase and identical charge mobility for ions and electrons. The Box model also assumes that electron diffusion and ion mobility are negligible in liquid argon during recombination.
In these models, the fraction of ionization electrons surviving recombination is a function of the number of ion-electron pairs per unit length, the electric field, the average ion-electron separation distance after thermalization and the angle of the particle with respect to the direction of the electric field -- plus the diffusion coefficient in the Birks model. Given the stringent assumptions, it is perhaps  not surprising that these models are in accordance to data only in specific regimes: the Birks model is generally used to describe recombination for low dE/dx, the Box model for high dEdX.
In LArTPC, the ICARUS and ArgoNeut have measured recombination in \cite{Amoruso:2004dy} and \cite{1748-0221-8-08-P08005} respectively. Since LArIAT uses the refurbished ArgoNeut TPC and cryostat at the same electric field,  LArIAT currently corrects for recombination using the ArgoNeut measured recombination parameters in \cite{1748-0221-8-08-P08005}.

\subsubsection{Space Charge Effect}
Slow-moving positive argon ions created during ionization can build-up in LArTPC, causing the distortion of the
electric field within the detector. This effect, called  ``space charge effect" leads to a displacement in the reconstructed position of the signal ionization electrons. In surface LArTPCs the space charge effect is primarily due to the rate of ionization produced by cosmic rays which is slowly drifting in the chamber at all times. Surface LArTPC of the size of several meters are expected to be modestly impacted from the space charge effect, where charge build-up create anisotropy of the electric field magnitude of the order of  5\% at a drift field of 500 V/cm \cite{SpaceCharge}. The smallness of the LArIAT drift volume is such that  effect of  space charge on the electric field is expected to be even smaller. \textcolor{red}{CHIEDI A FLAVIO}

@@

\textbf{Deconvolution.} Induction and collection planes have different field responses, given the different nature of the signals on these planes: the wires on the induction planes see inductive signal of the drifting charge, while the wires on the collection planes see the current derived from the charge entering the conductor. Thus, signals on the induction plane are bi-polar pulse and signal on the collection plane are unipolar pulses, see Figure \ref{fig:SignalProc} panel a. The first step in signal processing is deconvolution, that is a series of off-line algorithms geared towards undoing the detector effects. The result of the deconvolution step is  the production of  a comparable set waveforms on all planes presenting unipolar, approximately gaussian-like pulses (Figure \ref{fig:SignalProc} panel b). Signal from all planes are treated on equal footage beyond this point. Some LArTPC apply noise filtering in the frequency domain just after the deconvolution to clean up wire cross talk. Since signals from the LArIAT TPC are extremely clean, noise filtering is not necessary.\\

>>>>

Historically, the active medium in ionization detectors has been in the gaseous form. Carlo Rubbia first proposed the use of a Liquid Argon TPC for a neutrino experiment, ICARUS \cite{Rubbia:1977zz}, in 1977.  Using nobles elements in the liquid form for neutrino detectors is advantageous for several reasons.  The density of liquids is $\sim$1000 times greater than gases, augmenting the number of targets for neutrino's interaction in the same volume, in a effort to balance the smallness of neutrino cross section. Since the energy loss of charged particle is proportional to the target material density, as shown in the Bethe-Block equation (eq. \ref{eq:BB}), the increased density reflects into a proportionally higher energy loss, enhancing the calorimetry capability of detectors with a liquid active medium. Additionally, the ionization energy of liquids is smaller than gasses by the order of tens of eV. Thus, at the passage of charged particles, liquids generally produce more ionization electrons than gases for the same deposited energy, forcing the particles to deposit more energy in a shorter range. The downside of using noble liquid elements in experiments is that they require expensive cryogenic systems to cool the gas until it transitions to its the liquid form.
The properties of liquid argon in comparison liquid xenon -- a popular choice for dark matter and neutrinoless double beta decay detectors -- are summarized in table \ref{tab:properties}.  Albeit xenon would be more desirable than argon given some superior properties such as lower ionization energy and higher density and light yield, argon relative abundance abates the cost of argon compared to xenon, making argon a more viable choice for the construction of ton  (and kilo-ton) scale neutrino detectors.

@@

LArTPCs are some times referred as to ``electronic" bubble-chambers, for the similarity in the tracking and energy resolution which is coupled with an electronic readout of the imaging information in LArTPCs. Compared to these historic detectors however, LArTPC bestow tridimensional tracking and a self triggering mechanism provided by the scintillation light in the liquid argon.  An event display of a $\nu_\mu$ CC interaction candidate in the MicroBooNE detector is shown in picture \ref{fig:NuEvd} to display the level of spatial details these detectors are capable of; the color scale of the image is proportional to the energy deposited, hinting to these calorimetry capabilities of the detectors.
\begin{figure}[hbpt]
\centering
\includegraphics[width=\textwidth]{Chapter-2/Images/run3469_subrun1064_event53223_col.pdf}
\caption{Event display of a $\nu_\mu$ CC interaction candidate in the MicroBooNE detector.}
\label{fig:NuEvd}
\end{figure}

@@

\begin{figure}[hbpt]
\centering
\includegraphics[width=\textwidth]{Chapter-2/Images/Cartoon.png}
\caption{A cartoonish sketch of a LArTPC.}
\label{fig:cartoon}
\end{figure}

To the bare bones, a LArTPC is a bulk of liquid argon sandwiched in a flat capacitor, equipped with a light collection system, as the cartoon in \ref{fig:cartoon} shows. A uniform electric field of the order of 500 V/cm is maintained constant between the faces of the capacitor. The anode is sensitive to ionization charge and it is usually made of two or more planes segmented into several hundreds parallel sense wires a few millimeters apart; different geometries for the anode segmentation are under study \cite{1748-0221-8-07-P07002}.

Argon ionization and scintillation are the processes leveraged to detect particles in the LArTPC active volume.  When a ionizing radiation traverses the argon active volume it leaves a trail of ionization electrons along its trajectory and it excites the argon producing scintillation light -- details on the production and detection of ionization charge and scintillation light are provided in \ref{sec:light} and \ref{sec:light} respectively. The optical detector sees the argon scintillation light in matters of nanoseconds. This flash of light determines the start time of an event in the chamber, $t_0$. The uniform electric field drifts the ionization  electrons from the production point towards the anode in order of hundreds of microseconds or more depending on the chamber dimensions\footnote{The ionized argon also drifts, but  in the opposite directions compared to the electrons. Since the drift time is proportional to the particle mass,  the ions' drift time is much longer than the electrons'.  Ionized argon is collected on the cathode which is not instrumented, so it is not used to infer information about the interactions in the chamber.}. The anode sense wires see either an induced current by the drifting ionization charge (on induction planes) or an injection of such charge (collection plane).    An appropriate choice of the voltage bias on each wire plane assures ideal charge transparency, so that all the ionization charge is collected on the collection plane and none on the induction planes.

@@

Since the charge deposited by the ionizing radiation is proportional to the deposited energy and the charge collected on the sense plane is a function of the deposited charge, LArTPCs allow the measurement of the energy deposit in the active volume. Effects due to the presence of free charge and impurities in the active volume, such as a finite electron lifetime, recombination and space charge, complicate the relationship between deposited and collected charge affecting the measurement of the particle's energy, as described in the next section.

@@

Excitation and ionization of the detector medium occur in similar amounts. Since the ionizing collisions occur randomly, we can parametrize their number $k$ in a segment of length $s$ along the track  with a Poissonian function
\begin{equation}
P(k) = \frac{s^k}{k! \lambda^k} e^{-s/ \lambda }, 
\end{equation}
where $\lambda = 1/N_{e}\sigma_i$, with $N_{e}$ being the electron density of $\sigma_i$ the ionization cross-section per electron.  About 66\% of the ionizing collisions in Argon produce only a single electron/ion pair \cite{0034-4885-73-11-116201}; in the other cases, the transferred kinetic energy is enough for the primary electron to liberate one or more secondary electrons, which usually stay close to the original pair.  
Occasionally, electrons can receive enough energy to be ejected with high energy, forming a so-called ``$\delta$-ray": a detectable  short track off the particle trajectory, as shown in figure \ref{fig:delta}. 
The average number of $\delta$-ray  with energy E$>$E$_0$ per cm follows the empirical form
\begin{equation}
P(E>E_0) \sim \frac{y}{\beta^2 E_0},
\end{equation}
where $y$ is an empirical factor depending on the medium (0.114 for  gaseous Ar), and $\beta$ is $v/c$.

@@

\subsubsection{Purity \& Electron Life Time }
The presence of electronegative contaminants in liquid argon, such as oxygen $O_2$ and water $H_2O$, is particularly
pernicious, since these molecules quench the charge produced by the ionizing radiation.  Thus, amount of charge per unit of length $dQ/dx$ collected on the collection plane depends on the charge's production point in the detector: ionization produced  close to the cathode will see more impurities along its journey to the collection plane than ionization produced close to the anode, resulting in greater attenuation of its charge. As a result,  the amount of charge collected on the sense wires as a function of the traveled distance follows an exponential decay trend. The traveled distance is generally measured in terms of drift time and the  characteristic time constant of the exponential decay is called electron lifetime $\tau_e$. Figure \ref{fig:Elifetime} shows the typical life time for LArIAT data. The procedure to measure the electron lifetime in LArIAT is outlined in \cite{LArIATLifeTime}. LArIAT small drift distance (47 cm) allows for a relatively short electron life time. The life time for bigger detectors such as MicroBooNE, whose drift distance is 2.6 m, needs to be of the order of tens of milliseconds to allow a charge collection usable for physics analyses. Energy reconstruction in LArTPC applies a correction for the finite lifetime to calibrate the detector calorimetric response; details for LArIAT are provided in Section \ref{ch:energyCalibration}.

@@

Theoretical descriptions of recombination based on the Birks model and the Box model are provided in \cite{0370-1298-64-10-303} and  \cite{PhysRevA.36.614}, respectively. The Birks model assumes a gaussian spatial distribution around the particle trajectory during the entire recombination phase and identical charge mobility for ions and electrons. The Box model also assumes that electron diffusion and ion mobility are negligible in liquid argon during recombination.
In these models, the fraction of ionization electrons surviving recombination is a function of the number of ion-electron pairs per unit length, the electric field, the average ion-electron separation distance after thermalization and the angle of the particle with respect to the direction of the electric field -- plus the diffusion coefficient in the Birks model. Given the stringent assumptions, it is perhaps  not surprising that these models are in accordance to data only in specific regimes: the Birks model is generally used to describe recombination for low dE/dx, the Box model for high dE/dX.
In LArTPC, the ICARUS and ArgoNeut experiments have measured recombination in \cite{Amoruso:2004dy} and \cite{1748-0221-8-08-P08005} respectively. Since LArIAT uses the refurbished ArgoNeut TPC and cryostat at the same electric field,  LArIAT currently corrects for recombination using the ArgoNeut measured recombination parameters in \cite{1748-0221-8-08-P08005}.

\subsubsection{Space Charge Effect}
Slow-moving positive argon ions created during ionization can build-up in LArTPC, causing the distortion of the
electric field within the detector. This effect, called  ``space charge effect" leads to a displacement in the reconstructed position of the signal ionization electrons. In surface LArTPCs the space charge effect is primarily due to the rate of ionization produced by cosmic rays which is slowly drifting in the chamber at all times. Surface LArTPC of the size of several meters are expected to be modestly impacted from the space charge effect, where charge build-up create anisotropy of the electric field magnitude of the order of  5\% at a drift field of 500 V/cm \cite{SpaceCharge}. The smallness of the LArIAT drift volume is such that  effect of  space charge on the electric field is expected to be negligible. \textcolor{red}{CHIEDI A FLAVIO}

@@

\textbf{Deconvolution.} Induction and collection planes have different field responses, given the different nature of the signals on these planes: the wires on the induction planes see the inductive signal of the drifting charge, while the wires on the collection planes see the current derived from the charge entering the conductor. Thus, signals on the induction plane are bi-polar pulse and signal on the collection plane are unipolar pulses, see Figure \ref{fig:SignalProc} panel a. The first step in signal processing is deconvolution, that is a series of off-line algorithms geared towards undoing the detector effects. The result of the deconvolution step is  the production of  a comparable set waveforms on all planes presenting unipolar, approximately gaussian-like pulses (Figure \ref{fig:SignalProc} panel b). Signal from all planes are treated on equal footage beyond this point. Some LArTPC apply noise filtering in the frequency domain just after the deconvolution to clean up wire cross talk. Since signals from the LArIAT TPC are extremely clean, noise filtering is not necessary.\\
